Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
paper: some minor fixes and todos added

\subsection{Imputation placement}
\label{sec:placement}
Imputation operators must be placed so that no relational operator receives a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.

@@

\begin{definition}
Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p')$. 
\end{definition}

@@

For each sub-plan, we keep track of the estimated cardinality by using the histograms associated with
that plan. We use this estimate to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

@@

\srm{I find this table hard to follow -- I think it'd be good to explain one or two of the rows in detail, and to explain what the set L contains.}
Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing \nullv{} fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

@@

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes, each of which maintains its own Pareto frontier. 
Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Loss}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Loss}, \textsc{Time} \rangle$
as a total order, allowing us to appropriately select the frontier. Meanwhile, plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier corresponds to the collection of plans associate with the empty dirty set.

To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop. The minimal imputation (resp.\ drop) only imputes (resp.\ drops) the columns required by the relational operator immediately following the imputation. The maximal imputation imputes all of the attributes in the relation which are used by other operators in the query plan or are visible in the query output.

@@

The join optimization is based on the algorithm used in System R~\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the dirty-set specific Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

@@

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
effort and without changes to the optimizer, simply by adding time and loss functions
(\Cref{sec:cost}) which describe the new algorithm. In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

@@

Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set.
So in many cases we can drop some of these intermediate plans generated at each operator.
The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.
Planning times are discussed in \Cref{sec:results}.

>>>>

\subsection{Imputation placement}
\label{sec:placement}
Imputation operators must be placed so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.

@@

\begin{definition}
Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p'^\alpha)$. 
\end{definition}

@@

We use these histogram estimates to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

@@

\srm{I find this table hard to follow -- I think it'd be good to explain one or two of the rows in detail, and to explain what the set L contains.}
Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). 
\todobox{fix this explanation, not inverseley proprtional to what is stated here}{Imputing \nullv{} fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with.}
Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

@@

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dirty-set specific dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan per dirty set, because maintaining a Pareto frontier as a plan progresses through the optimizer's pipeline
allows us to obtain the final set of plans that best tradeoff computation
cost and imputation quality. Details of the plan-cache semantics are shown in~\Cref{fig:semantics-plan-cache}.

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Loss}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Loss}, \textsc{Time} \rangle$
as a total order, allowing us to appropriately select the frontier. Meanwhile, plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier produced by the optimizer corresponds to the collection of plans associated with the empty dirty set.

@@

The join optimization is based on the algorithm used in System R~\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

@@

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
effort and without changes to the optimizer, simply by adding time and loss functions
(\Cref{sec:cost-model}) which describe the new algorithm. In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

@@

Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
Recall that the planner maintains different plans for different dirty sets, keeping only those plans that are not dominated by others.
So in many cases we can drop some of the intermediate plans generated at each operator.
The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.
Planning times are discussed in \Cref{sec:results}.

### 2017-03-01 paper/experiments.tex
paper: some minor fixes and todos added

\subsection{Data sets}\label{subsec:datasets}
We collected three data sets our experiments.
For all data sets, we selected a subset of the original attributes.
We also transformed all data values to an integer representation by enumerating strings and transforming floating point values into an appropriate range.

@@

\subsubsection{American Community Survey}
For our final experiment, we run a simple aggregate query over data from the American
Community Survey (ACS), a comprehensive survey conducted by the U.S.
Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
data kindly provided by the authors of~\cite{akande2015empirical}. Given that the data had
been cleaned, we artificially dirtied it by removing 40\% of the values uniformly at random.
The final dataset consists of 671,153 rows and 37 integer columns.

@@

\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
\ProjectName{} optimizing for performance($\alpha=1$), and imputing at the base tables followed
by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
faster than imputing on the base tables. We can get another order-of-magnitude speedup when
optimizing for performance. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
separately run 200 times in both settings and then results from the two approaches were
paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
within each iteration of a query, and we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general,
the SMAPE relative to the base-imputation approach are low in all cases --- between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations relative to results when imputing on base table. Queries optimized for quality ($\alpha=0$) generally achieve
    lower error than queries optimized for efficiency ($\alpha=1$).}
\label{table:smape}
\end{table}

@@

For every imputation inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are not known until runtime.
Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.

>>>>

\subsection{Data sets}\label{subsec:datasets}
We collected three data sets for our experiments.
For all data sets, we selected a subset of the original attributes.
We also transformed all data values to an integer representation by enumerating strings and transforming floating point values into an appropriate range.

@@

\subsubsection{American Community Survey}
For our final experiment, we run a simple aggregate query over data from the American
Community Survey (ACS), a comprehensive survey conducted by the U.S.
Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
data kindly provided by the authors of~\cite{akande2015empirical}. Given that the data had
been cleaned, we artificially dirtied it by replacing 40\% of the values uniformly at random with \nullv{} values.
The final dataset consists of 671,153 rows and 37 integer columns.

@@

\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
\ProjectName{} optimizing for performance ($\alpha=1$), and imputing at the base tables followed
by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
faster than imputing on the base tables. We can get another order-of-magnitude speedup when
optimizing for performance. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
separately run 200 times in both settings and then results from the two approaches were
paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
within each iteration of a query, and we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general,
the SMAPE relative to the base-imputation approach are low in all cases --- between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

\begin{table}
\todo{extend this table to have counts with new results}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations relative to results when imputing on base table. Queries optimized for quality ($\alpha=0$) generally achieve
    lower error than queries optimized for efficiency ($\alpha=1$).}
\label{table:smape}
\end{table}

@@

For every imputation inserted into the query plan, a new statistical model is instantiated and trained before being used for imputation.
Although it is tempting to further optimize query execution in \ProjectName{} by pre-training imputation models while the analyst begins to make queries,
this is not desirable because the specific rows of data, set of complete attributes, and set of attributes to impute are not known until runtime.

\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-03-01 paper/relatedwork.tex
paper: some minor fixes and todos added

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of \nullv{}. As early as 1973,~\cite{codd1973understanding}
provides a treatment of the semantics of \nullv{}. Multiple
papers have described various (at times conflicting) treatments
of nulls~\cite{grant1977null}. \ProjectName's main design invariant - no relational operator
sees missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle \nullv{} value semantics, while guaranteeing soundness (modulo
imputation strategy) of the query evaluation.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional histogram transformations to account
for the changing nature of missing values.

>>>>

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of \nullv{}. As early as 1973,~\cite{codd1973understanding}
provided a treatment of the semantics of \nullv{}. Multiple
papers have described various (at times conflicting) treatments
of nulls~\cite{grant1977null}. \ProjectName's main design invariant - no relational operator
sees missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle \nullv{} value semantics, while guaranteeing query evaluation soundness.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
uses histogram transformations to account for the changing nature of missing values.
